---
title: "Reviving oceans"
subtitle: "..."
author: "James Goldie, 360info"
date: "2022-05-20"
code-fold: true
theme: style/article.scss
---

Analysis of data from [Sea Around Us](https://www.seaaroundus.org).

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(pins)
library(skimr)
library(here)
```

The data from [Sea Around Us](https://github.com/UBC-CIC/Sea-Around-Us) is kept on an S3 bucket as a flat CSV. It's a bit hefty, so let's cache it in `/data/.cache`:

```{r}
#| label: import
data_cache <- here("data", ".cache")
dir.create(data_cache, showWarnings = FALSE)

"https://d3cu2src083cxg.cloudfront.net/rfmo_query_result.csv" %>%
  { board_url(c(sau = .), cache = data_cache) } %>%
  pin_download("sau") ->
sau

saudata <- read_csv(sau)
```

Let's get a basic feel for the data. I really like Observable's [Summary Table](https://observablehq.com/@observablehq/summary-table) for this kind of thing, but I don't want to pull a 200MB csv through `ojs_define`, so I might use [`skimr::skim`](https://docs.ropensci.org/skimr/reference/skim.html) instead:

```{r}
saudata %>% skim(.data_name = "SAU data")
```

So this is annual catch data (the `catch_sum` and the `real_value`), broken down by:

* country (`fishing_entity`)
* species
* sector type
  - Industrial: ???
  _ Subsistence: somebody fishing to feed _themself_
  _ Artisanal: "small-scale, low-tech, low-capital, shore-based fishing by individuals"
  _ Recreational: ???
* gear type



